---
title: "Association rule mining"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
## Data Loading and Initial Exploration:
library(tidyverse)
library(igraph)
library(arules)
library(arulesViz)
library(knitr)

# Load the data
groceries_df <- read.delim("groceries.txt", header = FALSE, sep = ",", na.strings = c("", NaN))
groceries <- read.transactions("groceries.txt", format = "basket", sep = ",")
```


**Exploring the Dataset**:
    - displayed the first few rows of the dataset using `kable` to understand the structure and content of the data.

```{r echo = FALSE}
kable(head(groceries_df), format = "markdown")
```


**Top 20 Purchased Items**:
    - calculated the frequency of each item purchased and visualized the top 20 items with a horizontal bar chart.
    This helped me identify the most commonly purchased items.

```{r echo = FALSE}
# Generate item frequency table
item_freq <- itemFrequency(groceries, type = "absolute")

# Convert item frequency to a data frame
item_freq_df <- data.frame(item = names(item_freq), frequency = item_freq)

# Sort by frequency in descending order and keep top N items
top_n_items <- 20
top_items <- head(item_freq_df[order(-item_freq_df$frequency), ], top_n_items)

# Create a horizontal bar plot using ggplot2
ggplot(top_items, aes(x = frequency, y = reorder(item, frequency))) +
  geom_bar(stat = "identity", fill = "#1f77b4") +
  ggtitle("Top Purchased Items in Groceries") +
  xlab("Occurrences") +
  ylab("Purchased Item") +
  theme_minimal() +
  theme(axis.text.y = element_text(hjust = 0, size = 10))
```

# Applying the Association Rule


## Initial Rule: 
Initial Rule:
- I employed the apriori function to identify association rules in the dataset. My initial criteria were:
- Minimum support of 0.001: Meaning the items in the rule should appear in at least 0.1% of all transactions to be considered.
- Minimum confidence of 0.1: Meaning the rule should be correct 10% of the time.
- Minimum length of 2: I was only interested in rules that involve at least two items.
- The first five generated rules was displayed for a quick overview.

```{r include = FALSE}
rules <- apriori(groceries,parameter=list(support=0.001,confidence=0.1,minlen=2))
```

```{r echo = FALSE}
inspect(rules[1:5])
```
## visualize the distribution of rules based on their support and confidence.
##The shading by lift provided an indication of the "interestingness" of a rule.
```{r, echo=FALSE}
plot(rules, method = "scatter", measure = c("support", "confidence"), shading = "lift", jitter = 0)
```
There were too many rules so we pruned it down to make it simplier

## refine the rules to a minimum lift of 1 and a minimum confidence of 0.5. This filtered the rules to only those that are more significant and reliable.
```{r, echo=FALSE}
sub_rules = subset(rules, subset=lift > 1 & confidence > 0.5)
plot(subset(rules, subset=lift > 1 & confidence > 0.5), measure = c("support", "confidence"), shading = "lift",jitter = 0)
```

## with lift>1 and confidence>0.5, there are still too many rules and needed to be pruned further


4. **Final Analysis**:
    - I set stricter criteria for the rules:
        - Minimum confidence of 0.7: Indicating the rule should be correct 70% of the time.
        - Lift greater than 7: Suggesting that the items in the rule are bought together 7 times more often than expected if they were bought independently.


```{r, echo=FALSE}
better_rules = subset(rules, subset=lift > 7 & confidence > 0.7)

plot(better_rules, measure = c("support", "confidence"), shading = "lift", jitter = 0)
```
```{r, echo=FALSE}
plot(better_rules, method='graph')
#plot(better_rules, method='graph', engine='htmlwidget')
```
## The most significant rule can be summarized as: when liquor and red/blush wine are in the basket, bottled beer is also likely present with a confidence of approximately 90.5%.

## Summary:
- Through this analysis, I explored a grocery transaction dataset, visualized the most frequently purchased items, and utilized association rule mining to discern patterns.
- By adjusting parameters like support, confidence, and lift, I honed the rules to capture the most pertinent and reliable patterns.
- The graph I used at the end visually represents these rules, helping me quickly identify which items are frequently bought in tandem.